---
title: "Exploring the BRFSS data"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
```

### Load data

Make sure your data and R Markdown files are in the same directory. When loaded
your data file will be called `brfss2013`. Delete this note when before you submit 
your work. 

```{r load-data}
load("data/brfss2013.RData")
str(brfss2013)
```



* * *

## Part 1: Data
In 2013, 50 states, the District of Columbia, Guam, and Puerto Rico collected samples of interviews conducted both by landline telephone and cellular telephone.

Since 2011, BRFSS conducts both landline telephone- and cellular telephone-based surveys. In conducting the BRFSS landline telephone survey, interviewers collect data from a randomly selected adult in a household. In conducting the cellular telephone version of the BRFSS questionnaire, interviewers collect data from an adult who participates by using a cellular telephone and resides in a private residence or college housing.

In a telephone survey such as the BRFSS, a sample record is one telephone number in the list of all telephone numbers the system randomly selects for dialing. To meet the BRFSS standard for the participating states' sample designs, one must be able to justify sample records as a probability sample of all households with telephones in the state. All participating areas met this criterion in 2013.
Fifty-one projects used a disproportionate stratified sample (DSS) design for their landline samples. Guam and Puerto Rico used a simple random-sample design.

In the type of DSS design that states most commonly used in the BRFSS landline telephone sampling, BRFSS divides telephone numbers into two groups, or strata, which are sampled separately. The high-density and medium-density strata contain telephone numbers that are expected to belong mostly to households. Whether a telephone number goes into the high-density or medium-density stratum is determined by the number of listed household numbers in its hundred block, or set of 100 telephone numbers with the same area code, prefix, and first two digits of the suffix and all possible combinations of the last two digits. BRFSS puts numbers from hundred blocks with one or more listed household numbers (“1+ blocks,” or “banks”) in either the high-density stratum (“listed 1+ blocks”) or medium-density stratum (“unlisted 1 + blocks”). BRFSS samples the two strata to obtain a probability sample of all households with telephones. Cellular telephone sampling frames are commercially available and the system can call random samples of cellular telephone numbers, but doing so requires specific protocols. The basis of the 2013 BRFSS sampling frame is the Telecordia database of telephone exchanges (e.g., 617-492-0000 to 617-492-9999) and 1,000 banks (e.g., 617-492-0000 to 617-492-0999). The vendor uses dedicated cellular 1,000 banks, sorted on the basis of area code and exchange within a state. BRFSS forms an interval, K, by dividing the population count of telephone numbers in the frame, N, by the desired sample size, n. BRFSS divides the frame of telephone numbers into n intervals of size K telephone numbers. From each interval, BRFSS draws one 10-digit telephone number at random.

The target population for cellular telephone samples in 2013 consists of persons residing in a private residence or college housing, who have a working cellular telephone, are aged 18 and older, and received 90 percent or more of their calls on cellular telephones.

* * *

## Part 2: Research questions

**Research quesion 1:**
Is there a relationship between How Often Do You Drink Regular Soda Or Pop (ssbsugar) and (Ever Told) You Have Diabetes (diabete3)? And how about Had A Test For High Blood Sugar In Past Three Years (pdiabtst)

**Research quesion 2:**

genhlth: General Health

fruit: sum of below

* fruit1: How Many Times Did You Eat Fruit?
* fvbeans: How Many Times Did You Eat Beans Or Lentils?
* fvgreen: How Many Times Did You Eat Dark Green Vegetables?
* fvorang: How Many Times Did You Eat Orange-Colored Vegetables?
* vegetab1: How Many Times Did You Eat Other Vegetables?

How's the relation between general health (genhlth) and eating no fruit/vegi every day (fruit < 1)? 

Or form the data, if we know someone is not eating fruit, what is the probability he/she has health condition poor?


**Research quesion 3:**
Is Computed Height In Meters (htm4) following normal distribution?

* * *

## Part 3: Exploratory data analysis

NOTE: Insert code chunks as needed by clicking on the "Insert a new code chunk" 
button (green button with orange arrow) above. Make sure that your code is visible
in the project you submit. Delete this note when before you submit your work.

**Research quesion 1:**
Is there a relationship between How Often Do You Drink Regular Soda Or Pop (ssbsugar) and (Ever Told) You Have Diabetes (diabete3)? And how about Had A Test For High Blood Sugar In Past Three Years (pdiabtst)

Clean data first:
```{r}
brfss2013_clean = brfss2013[!rowSums(is.na(brfss2013[c('diabete3','ssbsugar')])), ]
str(brfss2013_clean$ssbsugar)
summary(brfss2013_clean$ssbsugar)
```
Transform data, so we only have "how often do you drink soda per day"
```{r}
brfss2013_clean$ssbsugar[(brfss2013_clean$ssbsugar>=100) & (brfss2013_clean$ssbsugar<200)] =
brfss2013_clean$ssbsugar[(brfss2013_clean$ssbsugar>=100) & (brfss2013_clean$ssbsugar<200)] - 100


brfss2013_clean$ssbsugar[(brfss2013_clean$ssbsugar>=200) & (brfss2013_clean$ssbsugar<300)] = 
(brfss2013_clean$ssbsugar[(brfss2013_clean$ssbsugar>=200) & (brfss2013_clean$ssbsugar<300)] - 200)/7

brfss2013_clean$ssbsugar[brfss2013_clean$ssbsugar>=300] = 
(brfss2013_clean$ssbsugar[brfss2013_clean$ssbsugar>=300] - 300)/30

brfss2013_clean$ssbsugar[brfss2013_clean$ssbsugar>=10] = 10

str(brfss2013_clean$ssbsugar)
summary(brfss2013_clean$ssbsugar)
```

Check transformed ssbsugar values:
```{r}
hist(brfss2013_clean$ssbsugar)
```

```{r}
brfss2013_clean$sugarH = brfss2013_clean$ssbsugar >2

table(brfss2013_clean$diabete3, brfss2013_clean$sugarH)
```

```{r}
table(brfss2013_clean$pdiabtst, brfss2013_clean$sugarH)
```

Conclusion for question 1:

There is no clear relationship between How Often Do You Drink Regular Soda Or Pop (ssbsugar) and (Ever Told) You Have Diabetes (diabete3).
And there no clear relationship between How Often Do You Drink Regular Soda Or Pop (ssbsugar) and Had A Test For High Blood Sugar In Past Three Years (pdiabtst).

* * *

**Research quesion 2:**

How's the relation between general health (genhlth) and eating no fruit/vegi every day (fruit < 1)? 

Or form the data, if we know someone is not eating fruit, what is the probability he/she has health condition poor?

Clean data first:
```{r}
brfss2013_clean = brfss2013[!rowSums(is.na(brfss2013[c('genhlth','fruit1', 'fvbeans', 'fvgreen', 'fvorang', 'vegetab1')])), ]
str(brfss2013_clean$genhlth)
```

Transform data, so that **fruit** represents how many time per day of all kind of vegi and fruit taken.

```{r}
brfss2013_clean$fruit1[(brfss2013_clean$fruit1>=100) & (brfss2013_clean$fruit1<200)] =
brfss2013_clean$fruit1[(brfss2013_clean$fruit1>=100) & (brfss2013_clean$fruit1<200)] - 100

brfss2013_clean$fruit1[(brfss2013_clean$fruit1>=200) & (brfss2013_clean$fruit1<300)] = 
(brfss2013_clean$fruit1[(brfss2013_clean$fruit1>=200) & (brfss2013_clean$fruit1<300)] - 200)/7

brfss2013_clean$fruit1[brfss2013_clean$fruit1>=300] = 
(brfss2013_clean$fruit1[brfss2013_clean$fruit1>=300] - 300)/30

brfss2013_clean$fruit1[brfss2013_clean$fruit1>=10] = 10

##
brfss2013_clean$fvbeans[(brfss2013_clean$fvbeans>=100) & (brfss2013_clean$fvbeans<200)] =
brfss2013_clean$fvbeans[(brfss2013_clean$fvbeans>=100) & (brfss2013_clean$fvbeans<200)] - 100

brfss2013_clean$fvbeans[(brfss2013_clean$fvbeans>=200) & (brfss2013_clean$fvbeans<300)] = 
(brfss2013_clean$fvbeans[(brfss2013_clean$fvbeans>=200) & (brfss2013_clean$fvbeans<300)] - 200)/7

brfss2013_clean$fvbeans[brfss2013_clean$fvbeans>=300] = 
(brfss2013_clean$fvbeans[brfss2013_clean$fvbeans>=300] - 300)/30

brfss2013_clean$fvbeans[brfss2013_clean$fvbeans>=10] = 10


##
brfss2013_clean$fvgreen[(brfss2013_clean$fvgreen>=100) & (brfss2013_clean$fvgreen<200)] =
brfss2013_clean$fvgreen[(brfss2013_clean$fvgreen>=100) & (brfss2013_clean$fvgreen<200)] - 100

brfss2013_clean$fvgreen[(brfss2013_clean$fvgreen>=200) & (brfss2013_clean$fvgreen<300)] = 
(brfss2013_clean$fvgreen[(brfss2013_clean$fvgreen>=200) & (brfss2013_clean$fvgreen<300)] - 200)/7

brfss2013_clean$fvgreen[brfss2013_clean$fvgreen>=300] = 
(brfss2013_clean$fvgreen[brfss2013_clean$fvgreen>=300] - 300)/30

brfss2013_clean$fvgreen[brfss2013_clean$fvgreen>=10] = 10


##
brfss2013_clean$fvorang[(brfss2013_clean$fvorang>=100) & (brfss2013_clean$fvorang<200)] =
brfss2013_clean$fvorang[(brfss2013_clean$fvorang>=100) & (brfss2013_clean$fvorang<200)] - 100

brfss2013_clean$fvorang[(brfss2013_clean$fvorang>=200) & (brfss2013_clean$fvorang<300)] = 
(brfss2013_clean$fvorang[(brfss2013_clean$fvorang>=200) & (brfss2013_clean$fvorang<300)] - 200)/7

brfss2013_clean$fvorang[brfss2013_clean$fvorang>=300] = 
(brfss2013_clean$fvorang[brfss2013_clean$fvorang>=300] - 300)/30

brfss2013_clean$fvorang[brfss2013_clean$fvorang>=10] = 10


##
brfss2013_clean$vegetab1[(brfss2013_clean$vegetab1>=100) & (brfss2013_clean$vegetab1<200)] =
brfss2013_clean$vegetab1[(brfss2013_clean$vegetab1>=100) & (brfss2013_clean$vegetab1<200)] - 100

brfss2013_clean$vegetab1[(brfss2013_clean$vegetab1>=200) & (brfss2013_clean$vegetab1<300)] = 
(brfss2013_clean$vegetab1[(brfss2013_clean$vegetab1>=200) & (brfss2013_clean$vegetab1<300)] - 200)/7

brfss2013_clean$vegetab1[brfss2013_clean$vegetab1>=300] = 
(brfss2013_clean$vegetab1[brfss2013_clean$vegetab1>=300] - 300)/30

brfss2013_clean$vegetab1[brfss2013_clean$vegetab1>=10] = 10

brfss2013_clean$fruit = 
    brfss2013_clean$fruit1 + 
    brfss2013_clean$fvbeans + 
    brfss2013_clean$fvgreen + 
    brfss2013_clean$fvorang + 
    brfss2013_clean$vegetab1
```

```{r}
hist(brfss2013_clean$fruit)

```
We define health condition above 'Good' to be 'Good', and blow 'Good' to be 'Poor'
```{r}
brfss2013_clean$genhlth[brfss2013_clean$genhlth=='Excellent'] = 'Good'
brfss2013_clean$genhlth[brfss2013_clean$genhlth=='Very good'] = 'Good'
brfss2013_clean$genhlth[brfss2013_clean$genhlth=='Fair'] = 'Poor'

table(brfss2013_clean$fruit < 1, brfss2013_clean$genhlth)

```

```{r}
# Percentage of Poor is:
p_poor = (69365 + 12859) / sum(table(brfss2013_clean$fruit < 1, brfss2013_clean$genhlth))
p_poor
```

```{r}
# Under poor health people, percentage of eating no vegi (brfss2013_clean$fruit < 1)
p_noVegi_knowing_poor = 12859 / (12859 + 69365)
p_noVegi_knowing_poor
```

```{r}
# Under good health people (> poor), percentage of eating no vegi (brfss2013_clean$fruit < 1)
p_noVegi_knowing_good = (30230) / (30230 + 325584)
p_noVegi_knowing_good
```

```{r}
# if we know someone is not eating fruit, what is the probability he/she has health condition poor?
# using Bayes' theorem:
p_noVegi_knowing_poor * p_poor / (p_noVegi_knowing_poor * p_poor + (1-p_poor)*p_noVegi_knowing_good)

```

```{r}
# Or direct calculation from data:
12859/(12859+30230)
```

```{r}
# if we know someone eating fruit, what is the probability he/she has health condition poor?# 
69365 / (325584+69365)
```

Thus: 
if we know someone is not eating fruit/vegi, the probability he/she being poor heath is around 30%.
if we know someone is eating fruit/vegi at least once a day, the probability he/she being poor heath is around 17.6%.
This means eating vegi/fruit is good for heath.

* * * 

**Research quesion 3:**
Is Computed Height In Meters (htm4) following normal distribution?

Clean data, remove NA:
```{r}
brfss2013 = brfss2013[!rowSums(is.na(brfss2013[c('htm4')])), ]
summary(brfss2013$htm4)
str(brfss2013$htm4)
```

Remove error data / outliers:
```{r}
brfss2013$htm4[brfss2013$htm4>244] = NA
brfss2013$htm4[brfss2013$htm4<100] = NA
summary(brfss2013$htm4)
hist(brfss2013$htm4)
```

Take 100 random samples out of the height and see if they are follow normal distribution
```{r}
height_sample <- brfss2013[sample(1:nrow(brfss2013), 100, replace=FALSE),]

qqnorm(height_sample$htm4)
qqline(height_sample$htm4)
```

Computed Height In Meters (htm4) is very closely following normal distribution.


